Future challenges for Research Software Engineering

### 2025-05-14 future_challenges_for_RSEng.tex
change What is boxes

\newcounter{whatiscounter}
\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{What is 2035}{
  enhanced,
  segmentation engine=empty,
%  beamer,
fuzzy shadow={2mm}{-1mm}{0mm}{0.1mm}%
{black!50!white},
  sharp corners,
  attach boxed title to top left={
    yshifttext=-1mm
  },
%  halign lower=flush center,
  colback=white,
  colframe=BurntOrange!75!white,
  fonttitle=\bfseries,
  attach boxed title to top left={yshift=-2mm,xshift=3mm},
  fontlower=\itshape,
  boxed title style={
    sharp corners,
    size=small,
    colback=BurntOrange!75!black,
    colframe=BurntOrange!75!black,
  } 
}{ist}

@@

\subsection{Definition and Evolution of RSEs}
\begin{whatis}{What do we want to have achieved}{definition-evolution-rse}
The term RSE is established in the minds of university boards, as a valuable member in the academic framework.
RSEs are developed using specialized master's program, and their support is valued by researchers. We still teach researchers basic software skills, but the content of this lecture has changed drastically.
\end{whatis}
% Here we consider how the topics of RSEng will change in the face of the digitalisation of society.
% Also here: What will our topics be after version control doesn't need to be taught anymore?
% Digitalisation will also feature more heavily in the domain curricula.
In this discussions the workshop participants found that
the digital skills of domain scientists will increase,
and software will become easier to interact with,
reducing reliance on RSEs to train newcomers on the most basic tasks
(JN: for example, versioning and archiving?)
(GC: Higher-level and more intuitive workflows that require less training).
User training sessions might focus on more advanced topics
that cater to a smaller audience, or involve mentoring more frequently.

@@

\subsection{Development of the RSE community}
\begin{whatis}{What do we want to have achieved}{development-rse-community}
National RSE communities are increasingly collaborating with local partners, fostering stronger networks across countries. Regular international conferences further amplify these connections, promoting the global exchange of ideas. The introduction of RSE Master's programs, coupled with a shift in culture towards valuing Research Software, has significantly elevated its profile. This growing recognition has led to a surge in membership within national societies, reflecting the increasing importance of RSEng in the research landscape.
\end{whatis}
In recent years, we saw the formation of RSE institutes and advocacy groups, such as
the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
the US Research Software Sustainability Institute\footnote{\url{https://urssi.us}},
the European Virtual Institute for Research Software Excellence\footnote{\url{https://everse.software}},
and the EOSC Opportunity Area Expert Group 7 on Research Software%
\footnote{\url{https://eosc.eu/opportunity-area-exp/oa7-research-software}}.
We expect these initiatives to help increase the visibility and accelerate
the professionalisation%
\footnote{MM: do we want to decide for UK or US English? Or keep it mixed? Not being a native English speaker I don't know how it reads.}
of the RSE role.
In addition, a multitude of RSE communities self-organised at the national
and regional level to support more viable RSE career paths in academia.
We can expect RSEs to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
The community of RSE's will get a significant boost in their acceptance if RSE Masters programmes
become available at various universities.
With increased numbers the visibility of the RSE community also to industry will become higher.
\begin{story}{Development of RSE community}{RSEcommunity}
Kay stood proudly at the podium of the first International Research Software Engineering Conference held in the Southern Hemisphere, a historic moment for both her career and the global RSE community. Having moved to South Africa, she spoke passionately about the unique challenges researchers in the Global South face—limited access to high-performance computing, unreliable funding, and the persistent digital divide. Her keynote highlighted how local innovation thrives despite constraints, often out of necessity rather than choice. The audience, a diverse mix of global experts, listened intently as she called for equitable collaborations, not just inclusion as an afterthought.
\end{story}

\subsection{Impact of AI and automation}
\begin{whatis}{What do we want to have achieved}{impact-ai-automation}
RSEs are educated about AI models, and can help researchers reap their benefits, while
clearly knowing their limits.
\end{whatis}
AI is already transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would be impossible or prohibitively time-consuming for humans alone to find.
In research and development, AI accelerates scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modeling and predictive tasks.
For software development, AI-powered tools automate code generation, bug detection, and testing, significantly increasing productivity and reducing human error.
Developers are shifting from manual coding to orchestrating AI-driven workflows, focusing more on high-level design, integration, and creative problem-solving while AI handles repetitive or routine tasks.
The integration of AI is democratizing access to advanced technologies and analytics, enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
As AI continues to evolve, it will further blur the lines between human and machine roles, making research and software development faster, more efficient, and more innovative than ever before.
With the general availability of AI the interaction with researchers will change at a fundamental level.
Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to it,
It can be foreseen, that AI drives this low-key interaction to a highly specialized level of knowledge,
since AI copilots are available 24/7.
Since AI systems are so democratically available and will be used as black boxes,
the interpretation of their outputs will be far more imortant.
Currently we are expanding and documenting our work in the open and making our code available,
therefore there will be more structured material to train GenAI on.
However, as people get quicker answers from GenAI, the community resources (\eg Stack Overflow, Wikipedia)
get poorer.

@@

\subsection{Impact of Complexity and Usability in Research Software}
\begin{whatis}{What do we want to have achieved}{abstraction}
With a palette of readily reusable software components,
and with their skills in managing hierarchical systems of complex software components with clear interfaces,
RSEs can now build systems of scale and capability not previously possible.
\end{whatis}

@@

\subsection{Ethics and social considerations}
\begin{whatis}{What do we want to have achieved}{ethics}
The principles of Good Scientific Practice \cite{dfg_gsp} have been a cornerstone in the conduct of research.
But their relevance hinges on constant updates in light of how research is performed.
By 2035 we want to have achieved accepted rules on the use of AI in research tasks and software development.
Open Science and Open Research are well-established in academia, and supported by respective Open Research Software
strategies of the political actors. We also hope, that Data Awareness and Data Protection are also still values
that are protected.
\end{whatis}

@@

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{What do we want to have achieved}{academic}
We expect research software to be a central part of scientific practice on the same level as data and publications.
This will be supported by the first generations of RSE-Master graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, that are known by 2035 in university boards, we expect the concept of RSE units to be in the minds
of academic staff at any level.
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects \cite{Kempf2025-draft}.
We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
At universities which have them, we want to see that researchers naturally accept them as part of their used services.
Additionally researchers will seamlessly interact with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative (cite WissZeitVG here), has organisationally often adopted time-scales for change that are more reminiscient to civil services like pension funds.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref} \cite{hiddenref} hints at the potential for specialisation of research tasks. This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
We have already seen some specialisation trends in recent years such as
the HPC-RSE and AI-RSE which are quite recent developments, whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
But we could imagine new RSE specialisations to emerge in the fields
of quantum computing, such as the Quantum-RSE, and in federated online platforms for research.
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

@@

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
\begin{whatis}{What do we want to have achieved}{career-paths}
We would like to see the emergence of more dedicated personel, i.e. other RTPs next to the RSE.
Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg in think tanks, NGOs,
or organisations that do knowledge transfer).
\end{whatis}
RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units.\cite{Kempf2025-draft}
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding is more difficult to acquire in lower levels of an organisation.
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg, Göttingen).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups themselves prove themselves valuable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.

>>>>

\newcounter{whatiscounter}
\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{What do we want to have achieved}{
  enhanced,
  segmentation engine=empty,
%  beamer,
fuzzy shadow={2mm}{-1mm}{0mm}{0.1mm}%
{black!50!white},
  sharp corners,
  attach boxed title to top left={
    yshifttext=-1mm
  },
%  halign lower=flush center,
  colback=white,
  colframe=BurntOrange!75!white,
  fonttitle=\bfseries,
  attach boxed title to top left={yshift=-2mm,xshift=3mm},
  fontlower=\itshape,
  boxed title style={
    sharp corners,
    size=small,
    colback=BurntOrange!75!black,
    colframe=BurntOrange!75!black,
  } 
}{ist}

@@

\subsection{Definition and Evolution of RSEs}
\begin{whatis}{}{definition-evolution-rse}
The term RSE is established in the minds of university boards, as a valuable member in the academic framework.
RSEs are developed using specialized master's program, and their support is valued by researchers. We still teach researchers basic software skills, but the content of this lecture has changed drastically.
\end{whatis}
% Here we consider how the topics of RSEng will change in the face of the digitalisation of society.
% Also here: What will our topics be after version control doesn't need to be taught anymore?
% Digitalisation will also feature more heavily in the domain curricula.
In this discussions the workshop participants found that
the digital skills of domain scientists will increase,
and software will become easier to interact with,
reducing reliance on RSEs to train newcomers on the most basic tasks
(JN: for example, versioning and archiving?)
(GC: Higher-level and more intuitive workflows that require less training).
User training sessions might focus on more advanced topics
that cater to a smaller audience, or involve mentoring more frequently.

@@

\subsection{Development of the RSE community}
\begin{whatis}{}{development-rse-community}
National RSE communities are increasingly collaborating with local partners, fostering stronger networks across countries. Regular international conferences further amplify these connections, promoting the global exchange of ideas. The introduction of RSE Master's programs, coupled with a shift in culture towards valuing Research Software, has significantly elevated its profile. This growing recognition has led to a surge in membership within national societies, reflecting the increasing importance of RSEng in the research landscape.
\end{whatis}
In recent years, we saw the formation of RSE institutes and advocacy groups, such as
the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
the US Research Software Sustainability Institute\footnote{\url{https://urssi.us}},
the European Virtual Institute for Research Software Excellence\footnote{\url{https://everse.software}},
and the EOSC Opportunity Area Expert Group 7 on Research Software%
\footnote{\url{https://eosc.eu/opportunity-area-exp/oa7-research-software}}.
We expect these initiatives to help increase the visibility and accelerate
the professionalisation%
\footnote{MM: do we want to decide for UK or US English? Or keep it mixed? Not being a native English speaker I don't know how it reads.}
of the RSE role.
In addition, a multitude of RSE communities self-organised at the national
and regional level to support more viable RSE career paths in academia.
We can expect RSEs to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
The community of RSE's will get a significant boost in their acceptance if RSE Masters programmes
become available at various universities.
With increased numbers the visibility of the RSE community also to industry will become higher.
\begin{story}{Development of RSE community}{RSEcommunity}
Kay stood proudly at the podium of the first International Research Software Engineering Conference held in the Southern Hemisphere, a historic moment for both her career and the global RSE community. Having moved to South Africa, she spoke passionately about the unique challenges researchers in the Global South face—limited access to high-performance computing, unreliable funding, and the persistent digital divide. Her keynote highlighted how local innovation thrives despite constraints, often out of necessity rather than choice. The audience, a diverse mix of global experts, listened intently as she called for equitable collaborations, not just inclusion as an afterthought.
\end{story}

\subsection{Impact of AI and automation}
\begin{whatis}{}{impact-ai-automation}
RSEs are educated about AI models, and can help researchers reap their benefits, while
clearly knowing their limits.
\end{whatis}
AI is already transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would be impossible or prohibitively time-consuming for humans alone to find.
In research and development, AI accelerates scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modeling and predictive tasks.
For software development, AI-powered tools automate code generation, bug detection, and testing, significantly increasing productivity and reducing human error.
Developers are shifting from manual coding to orchestrating AI-driven workflows, focusing more on high-level design, integration, and creative problem-solving while AI handles repetitive or routine tasks.
The integration of AI is democratizing access to advanced technologies and analytics, enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
As AI continues to evolve, it will further blur the lines between human and machine roles, making research and software development faster, more efficient, and more innovative than ever before.
With the general availability of AI the interaction with researchers will change at a fundamental level.
Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to it,
It can be foreseen, that AI drives this low-key interaction to a highly specialized level of knowledge,
since AI copilots are available 24/7.
Since AI systems are so democratically available and will be used as black boxes,
the interpretation of their outputs will be far more imortant.
Currently we are expanding and documenting our work in the open and making our code available,
therefore there will be more structured material to train GenAI on.
However, as people get quicker answers from GenAI, the community resources (\eg Stack Overflow, Wikipedia)
get poorer.

@@

\subsection{Impact of Complexity and Usability in Research Software}
\begin{whatis}{}{abstraction}
With a palette of readily reusable software components,
and with their skills in managing hierarchical systems of complex software components with clear interfaces,
RSEs can now build systems of scale and capability not previously possible.
\end{whatis}

@@

\subsection{Ethics and social considerations}
\begin{whatis}{}{ethics}
The principles of Good Scientific Practice \cite{dfg_gsp} have been a cornerstone in the conduct of research.
But their relevance hinges on constant updates in light of how research is performed.
By 2035 we want to have achieved accepted rules on the use of AI in research tasks and software development.
Open Science and Open Research are well-established in academia, and supported by respective Open Research Software
strategies of the political actors. We also hope, that Data Awareness and Data Protection are also still values
that are protected.
\end{whatis}

@@

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{}{academic}
We expect research software to be a central part of scientific practice on the same level as data and publications.
This will be supported by the first generations of RSE-Master graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, that are known by 2035 in university boards, we expect the concept of RSE units to be in the minds
of academic staff at any level.
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects \cite{Kempf2025-draft}.
We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
At universities which have them, we want to see that researchers naturally accept them as part of their used services.
Additionally researchers will seamlessly interact with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative (cite WissZeitVG here), has organisationally often adopted time-scales for change that are more reminiscient to civil services like pension funds.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref} \cite{hiddenref} hints at the potential for specialisation of research tasks. This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
We have already seen some specialisation trends in recent years such as
the HPC-RSE and AI-RSE which are quite recent developments, whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
But we could imagine new RSE specialisations to emerge in the fields
of quantum computing, such as the Quantum-RSE, and in federated online platforms for research.
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

@@

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
\begin{whatis}{}{career-paths}
We would like to see the emergence of more dedicated personel, i.e. other RTPs next to the RSE.
Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg in think tanks, NGOs,
or organisations that do knowledge transfer).
\end{whatis}
RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units.\cite{Kempf2025-draft}
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding is more difficult to acquire in lower levels of an organisation.
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg, Göttingen).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups themselves prove themselves valuable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.
